---
title: "Project"
author: "Atharva Jadhav | Harshali Gaikwad | Sriujan Harihar"
date: "2025-04-12"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r}
library(tidyverse)
library(tseries)    
library(forecast)   
library(Metrics)
library(ggplot2)
library(tibble)
library(scales)
```

**Part I: Data Loading**

```{r}
data = read.csv("pollution_us_2000_2016.csv")
```

```{r}
data_df = data.frame(data)
```

**Part II: Data Cleaning**

```{r}
na_counts <- colSums(is.na(data_df)) # Find the number of NA values in each column
print(na_counts)
```

```{r}
cleaned_df <- data.frame(lapply(data_df, function(x) {
  if (is.numeric(x)) {
    x[is.na(x)] <- mean(x, na.rm = TRUE)
  }
  return(x)
}))

```

```{r}
na_counts <- colSums(is.na(cleaned_df))
print(na_counts)
```

**Part III: Time Series \~ County - Cook**

```{r}
cleaned_df$Date.Local <- as.POSIXct(cleaned_df$Date.Local, format = "%Y-%m-%d")
cleaned_df_cook <- subset(cleaned_df, County == "Cook")

cleaned_df_cook_floored <- cleaned_df_cook %>%
  mutate(YearMonth = floor_date(Date.Local, "month"))
```

```{r}
monthly_data <- cleaned_df_cook_floored %>% 
  group_by(YearMonth) %>%
  summarise(mean_CO = mean(CO.Mean, na.rm = TRUE))
```

**1. Plot the Time Series for Cook**

```{r}

monthly_data$YearMonth <- as.Date(monthly_data$YearMonth)

monthly_data_2004 <- monthly_data %>% 
  filter(format(YearMonth, "%Y") == "2004")

ggplot(monthly_data_2004, aes(x = YearMonth, y = mean_CO)) +
  geom_line(color = "darkgreen") +
  labs(title = "Monthly Average CO Values (Year 2004) Cook",
       x = "Month",
       y = "Mean CO") +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +  # Show each month name
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

**2. Plot the Monthly Mean Time Series for Cook**

```{r}
start_year <- year(min(monthly_data$YearMonth))
start_month <- month(min(monthly_data$YearMonth))
ts_CO <- ts(monthly_data$mean_CO, start = c(start_year, start_month), frequency = 12)

# Plot the time series
plot(ts_CO, main = "Time Series of Monthly Mean CO - Cook", ylab = "Mean CO", xlab = "Time")

```

**3. Check for Stationarity using Augmented Dickey-Fuller Test**

```{r}
# Perform the Augmented Dickey-Fuller Test for stationarity.
adf_test <- adf.test(ts_CO, alternative = "stationary")
print(adf_test)
```

The data from the output of the Dickey-Fuller Test states that the data we have is not stationary. We will take care of it in the ARIMA model by differencing.

**4. Find out the ACF and PACF**

```{r}
# Plot ACF and PACF for the series
Acf(ts_CO, main = "ACF Plot - Cook")
Pacf(ts_CO, main = "PACF Plot - Cook")

#Plot the seasonal graph for checking the seasonality. 
ggseasonplot(ts_CO, year.labels = TRUE)
```

Now using the above plot we see that there is seasonality for the year 2000 - 2004 and there is no seasonality seen for the rest of the years. Hence we split the time series fat 2004, and we take the years 2004 - 2013.

```{r}
ts_CO_partitioned <- window(ts_CO, start = c(2005, 1))
adf_test <- adf.test(ts_CO_partitioned, alternative = "stationary")
print(adf_test)

ts_CO_partitioned_diff <- diff(ts_CO_partitioned)
print(adf.test(ts_CO_partitioned_diff, alternative = "stationary"))

```

**5. Find out the ACF and PACF after differencing the data.**

```{r}
Acf(ts_CO_partitioned_diff, main = "ACF Plot of Differenced Partitioned Series- Cook")
Pacf(ts_CO_partitioned_diff, main = "PACF Plot of Differenced Partitioned Series-  Cook")

ggseasonplot(ts_CO_partitioned_diff,year.labels = TRUE, main = "Seasonal Subseries Plot")


boxplot(ts_CO_partitioned_diff ~ cycle(ts_CO_partitioned_diff), xlab = "Month", ylab = "Mean CO", main = "Seasonal Boxplot - Cook")
```

**6. ARMA Model fitting to the Time Series Data**

```{r}
auto_model <- auto.arima(ts_CO_partitioned_diff, seasonal = FALSE)
summary(auto_model)
checkresiduals(auto_model)
```

**7. Forecasting and Model Evaluation for Cook**

```{r}

future_forecast <- forecast(auto_model, h = 12)
autoplot(future_forecast)
summary(future_forecast)

```

**Part IV: Time Series \~ Maricopa**

```{r}

cleaned_df_maricopa <- subset(cleaned_df, County == "Maricopa")

cleaned_df_maricopa_floored <- cleaned_df_maricopa %>%
  mutate(YearMonth = floor_date(Date.Local, "month"))

```

```{r}

monthly_data_maricopa <- cleaned_df_maricopa_floored %>% 
  group_by(YearMonth) %>%
  summarise(mean_CO = mean(CO.Mean, na.rm = TRUE))

```

**1. Plot the Time Series for Maricopa**

```{r}

monthly_data_maricopa$YearMonth <- as.Date(monthly_data_maricopa$YearMonth)

monthly_data_maricopa_2004 <- monthly_data_maricopa %>% 
  filter(format(YearMonth, "%Y") == "2004")

ggplot(monthly_data_maricopa_2004, aes(x = YearMonth, y = mean_CO)) +
  geom_line(color = "darkgreen") +
  labs(title = "Monthly Average CO Values (Year 2004) Maricopa",
       x = "Month",
       y = "Mean CO") +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +  # Show each month name
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

**2. Plot the Monthly Mean Time Series for Maricopa**

```{r}
start_year_mc <- year(min(monthly_data_maricopa$YearMonth))
start_month_mc <- month(min(monthly_data_maricopa$YearMonth))
ts_CO_mc <- ts(monthly_data_maricopa$mean_CO, start = c(start_year_mc, start_month_mc), frequency = 12)

# Plot the time series
plot(ts_CO_mc, main = "Time Series of Monthly Mean CO Maricopa", ylab = "Mean CO", xlab = "Time")
```

**3. Check for Stationarity using Augmented Dickey-Fuller Test**

```{r}
# Perform the Augmented Dickey-Fuller Test for stationarity.

adf_test_mc <- adf.test(ts_CO_mc, alternative = "stationary")
print(adf_test_mc)
```

Since the data is already stationary according to the p-value which is 0.01, hence plotting now ACF and PACF

**4. Find out the ACF and PACF for Maricopa**

```{r}
# Plot ACF and PACF for the series

Acf(ts_CO_mc, main = "ACF Plot - Maricopa")
Pacf(ts_CO_mc, main = "PACF Plot - Maricopa")


```

**5. Checking for Seasonality of Time Series Data for Maricopa**

```{r}

ggseasonplot(ts_CO_mc, year.labels = TRUE)
```

Since the data is stationary and the data is seasonal we are using SARIMA model.

**6. SARIMA Model fitting to the Time Series Data for Maricopa**

```{r}
auto_model_mc <- auto.arima(ts_CO_mc, seasonal = TRUE)
summary(auto_model_mc)
checkresiduals(auto_model_mc)
```

**7. Forecasting and Model Evaluation for Maricopa**

```{r}

future_forecast_mc <- forecast(auto_model_mc, h = 12)
autoplot(future_forecast_mc)
summary(future_forecast_mc)

```

**Part V: Comparison ~ Cook and Maricopa (2000)**

```{r}
resid1 <- residuals(auto_model)
resid2 <- residuals(auto_model_mc)

min_len <- min(length(resid1), length(resid2))

# Trim both residuals to that length
resid1_trim <- resid1[1:min_len]
resid2_trim <- resid2[1:min_len]
time_vals <- time(resid1)[1:min_len]  # assuming both have same time index structure

# Now create the tibble
df_resid <- tibble::tibble(
  Time = time_vals,
  CountyCook = as.numeric(resid1_trim),
  CountyMaricopa = as.numeric(resid2_trim)
)

# Reshape for plotting
df_long <- tidyr::pivot_longer(df_resid, cols = c(CountyCook, CountyMaricopa),
                                names_to = "County", values_to = "Residual")

# Plot
library(ggplot2)
ggplot(df_long, aes(x = Time, y = Residual, color = County)) +
  geom_line() +
  labs(title = "Residuals Comparison by County",
       x = "Time", y = "Residuals") +
  theme_minimal()

df_density <- data.frame(
  Residual = c(resid1_trim, resid2_trim),
  County = rep(c("Cook County", "Maricopa County"), each = length(resid1_trim))
)

ggplot(df_density, aes(x = Residual, fill = County)) +
  geom_density(alpha = 0.5) +
  labs(title = "Residual Distribution by County") +
  theme_minimal()
```
